---
title: "Understanding  Missing Values in R"
subtitle: "Handling Missing Data in R"
author: "Adejumo Ridwan Suleiman"
date: today
format: 
  revealjs:
    theme: default
    transition: slide
    highlight-style: github
    code-line-numbers: true
    incremental: true
---

# Understanding Missing Data

## What is Missing Data? 📖

Missing data occurs when **no data value is stored** for a variable in an observation. In datasets, this appears as:

- `NA` (Not Available) in R
- Empty cells in spreadsheets  
- Blank responses in surveys
- System failures during data collection



# Types of Missing Data 🧩


## MCAR - Missing Completely at Random

- **Definition**: Missingness is unrelated to observed or unobserved data
- **Example**: Survey responses lost due to computer malfunction
- **Implication**: Safe to use listwise deletion

## MAR - Missing at Random  

- **Definition**: Missingness depends on observed data, but not unobserved data
- **Example**: Older participants more likely to skip technology questions
- **Implication**: Can be handled with multiple imputation

## MNAR - Missing Not at Random

- **Definition**: Missingness depends on the unobserved (missing) values themselves
- **Example**: High earners refusing to report income
- **Implication**: Requires specialized modeling approaches



## Impact on Analysis ⚠️

```{r}
#| echo: true
#| eval: false

# Original dataset: 1000 observations
# After listwise deletion: 300 observations
# Lost 30% of data!
```

### Consequences:

- **Reduced sample size** → Lower statistical power
- **Biased estimates** → Incorrect conclusions  
- **Invalid inferences** → Poor generalizability
- **Increased standard errors** → Wider confidence intervals



# Spotting Missing Data in R 🔍

## Basic Detection

```{r}
#| echo: true
#| eval: false

# Check for any missing values
any(is.na(data))

# Count missing values per column
colSums(is.na(data))

# Proportion of missing data
mean(is.na(data$variable))
```

## Visualization Tools

```{r}
#| echo: true
#| eval: false

# Using VIM package
library(VIM)
VIM::aggr(data, col = c('navyblue', 'red'))

# Using naniar package  
library(naniar)
naniar::vis_miss(data)
naniar::gg_miss_var(data)
```


